Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-18 survey/survey.tex
subsection 5.2 and citations

%%%
\subsection{Infinite Sets}

@@

%%%
\subsection{Topological Spaces}

@@

%%%
\subsection{Theseus}

The most elementary control structure in a programming language is the
ability to conditionally execute one of several possible code
fragments. Expressing such an abstraction using predicates and nested
\textbf{if}-expressions makes it difficult for both humans and
compilers to reason about the control flow structure. Instead, in
modern functional languages, this control flow paradigm is elegantly
expressed using \emph{pattern-matching}. This approach yields code
that is not only more concise and readable but also enables the
compiler to easily verify two crucial properties: (i) non-overlapping
patterns and (ii) exhaustive coverage of a datatype using a collection
of patterns. Indeed most compilers for functional languages perform
these checks, warning the user when they are violated. At a more
fundamental level, e.g., in type theories and proof assistants, these
properties are actually necessary for correct reasoning about
programs. Our first insight, explained in this section, is that these
properties, perhaps surprisingly, are sufficient to produce a simple
and intuitive first-order reversible programming language.

We start with a small illustrative example, written in a Haskell-like
syntax.  Fig.~\ref{fig:intro-ex-1} gives the skeleton of a function
\verb|f| that accepts a value of type \verb|Either Int| \verb|Int|; the
patterns on the left-hand side exhaustively cover every possible
incoming value and are non-overlapping. Similarly,
Fig.~\ref{fig:intro-ex-2} gives the skeleton for a function~\verb|g|
that accepts a value of type \verb|(Bool,Int)|; again the patterns on
the left-hand side exhaustively cover every possible incoming value
and are non-overlapping. Now we claim that since the types
\verb|Either Int Int| and \verb|(Bool,Int)| are isomorphic, we can
combine the patterns of \verb|f| and \verb|g| into \emph{symmetric
  pattern-matching clauses} to produce a reversible function between
the types \verb|Either Int Int| and
\verb|(Bool,Int)|. Fig.~\ref{fig:intro-ex-3} gives one such function;
there, we suggestively use \verb|<->| to indicate that the function
can be executed in either direction. This reversible function is
obtained by simply combining the non-overlapping exhaustive patterns
on the two sides of a clause. In order to be well-formed in either
direction, these clauses are subject to the constraint that each
variable occurring on one side must occur exactly once on the other
side (and with the same type). Thus it is acceptable to swap the
second and third right-hand sides of \verb|h| but not the first and
second ones.

%%%
\subsection{Quantum Computing}

Extension of Theseus in 2018 FoSSaCS paper.

Exclusive Disjunctions

Quantum over finite field

\jc{The reason I commented this out is that it is under-justified. The
reader will simply not understand what these next few lines are really
saying.}

Indeed one should take the physical principles underlying quantum
mechanics, the most successful physical theory known to us and adapt
computation to ``learn'' from these principles. To illustrate the
depth of our crisis, Scott Aaronson, Umesh Vazirani, and others have
proposed the following puzzle.

One of these wild claims must be true!:
\begin{itemize}
\item the extended Church-Turing thesis is false, or
\item quantum physics is false, or
\item there is an efficient classical algorithm for factoring
\end{itemize}
Indeed, if quantum physics is correct then there is an efficient quantum
algorithm for factoring (Shor). If there is no efficient classical algorithm
for factoring then the extended Church-Turing thesis is false.

\begin{verbatim}
check-password (guess) =
  if guess == "10"
  then True
  else False
\end{verbatim}

>>>>

\begin{verbatim}
check-password (guess) =
  guess == "10"
\end{verbatim}

%%%
\subsection{Theseus and Quantum Control}

The $\Pi$ family of languages semantically captures the principles of
reversibility and conservation of information. As a programming
language it has some mixed properties: small programs are relatively
easy to write; for some special classes of programs, it is even
possible to define a methodology to write large $\Pi$ programs,
including a self-interpreter for $\Pi$~\cite{isoint}. In general,
however, the point-free style of combinators used in $\Pi$ becomes
awkward and a new approach appears more suitable. To that end, we note
that $\Pi$ encodes the most elementary control structure in a
programming language-- which is the ability to conditionally execute
one of several possible code fragments-- using combinators. Expressing
such an abstraction using combinators or even predicates and nested
\textbf{if}-expressions makes it difficult for both humans and
compilers to write, understand, and reason about the control flow
structure of the program. Instead, in modern functional languages,
this control flow paradigm is elegantly expressed using
\emph{pattern-matching}. This approach yields code that is not only
more concise and readable but also enables the compiler to easily
verify two crucial properties: (i) non-overlapping patterns and (ii)
exhaustive coverage of a datatype using a collection of
patterns. Indeed most compilers for functional languages perform these
checks, warning the user when they are violated. At a more fundamental
level, e.g., in type theories and proof assistants, these properties
are actually necessary for correct reasoning about programs. Our
insight is that these properties, perhaps surprisingly, are sufficient
to produce a simple and intuitive first-order reversible programming
language which we call \emph{Theseus}.

We provide a small illustrative example, written in a Haskell-like
syntax.  Fig.~\ref{fig:intro-ex-1} gives the skeleton of a function
\verb|f| that accepts a value of type \verb|Either Int| \verb|Int|;
the patterns on the left-hand side exhaustively cover every possible
incoming value and are non-overlapping. Similarly,
Fig.~\ref{fig:intro-ex-2} gives the skeleton for a function~\verb|g|
that accepts a value of type \verb|(Bool,Int)|; again the patterns on
the left-hand side exhaustively cover every possible incoming value
and are non-overlapping. Now we claim that since the types
\verb|Either Int Int| and \verb|(Bool,Int)| are isomorphic, we can
combine the patterns of \verb|f| and \verb|g| into \emph{symmetric
  pattern-matching clauses} to produce a reversible function between
the types \verb|Either Int Int| and
\verb|(Bool,Int)|. Fig.~\ref{fig:intro-ex-3} gives one such function;
there, we suggestively use \verb|<->| to indicate that the function
can be executed in either direction. This reversible function is
obtained by simply combining the non-overlapping exhaustive patterns
on the two sides of a clause. In order to be well-formed in either
direction, these clauses are subject to the constraint that each
variable occurring on one side must occur exactly once on the other
side (and with the same type). Thus it is acceptable to swap the
second and third right-hand sides of \verb|h| but not the first and
second ones. With some additional work, it is possible to extend
Theseus to a full-fledged reversible programming
language~\cite{theseus}. With just one additional insight, Theseus can
be extended with superpositions and becomes a quantum programming
language~\cite{10.1007/978-3-319-89366-2_19}.

%%%
\subsection{Richer Data: Infinite Sets and Topological Spaces}

@@

@@

topological spaces have strange cardinalities: negative types
fractional types etc
